---
title: "wang_final"
author: "Binchen Mao"
date: "2016年6月24日"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Statistical analysis of Dr. Zhun Wang's dataset

## *load library*
```{r,message=FALSE}
rm(list=ls())
library("lme4")
library("multcomp")
library("lattice")
library(nlme)
```
## Data input and preprocessing
```{r}
wang.raw=read.csv("C:\\Users\\binchen.mao\\Documents\\crownbio\\For WangZhun\\data_06102016.csv",h=T)
days = c(0,4,7,11,14,18,21)
wang.long = reshape(idvar='Temp.ID',varying = c(3:9),timevar = 'PGday',times=days,data=wang.raw,direction = 'long',v.names = 'TV')
group=read.csv("C:\\Users\\binchen.mao\\Documents\\crownbio\\For WangZhun\\group_06102016.csv",h=T)
#group$Treatment = factor(group$Treatment,ordered = T, levels =c("Control","BA-087-mAB2","Anti-hCTLA4") )
wang = merge(wang.long,group)
wang[order(wang$mouse,wang$PGday),][1:10,]
```
##data analysis
###Check the average Tumor volume for each **Treatment** or **PGday**
```{r}
tapply(wang$TV,wang$Treatment,mean)
tapply(wang$TV,wang$PGday,mean)
```

###Making boxplot to check distribution of Tumor Volume
```{r}
plot(TV~Treatment,data=wang)
plot(log(TV+1)~Treatment,data=wang)
plot(TV**(1/3)~Treatment,data=wang)
```

###Checking baseline level TV for each treatment
```{r}
TV0 = wang[wang$PGday==0,]
tapply(TV0$TV,TV0$Treatment,mean)
tapply(TV0$TV,TV0$Treatment,sd)
TV0 = TV0[,c('mouse','TV')]
names(TV0)[2]='TV0'
```

###Run Two way RM ANOVA and posthoc test (Tukey's HSD),
```{r,warning=FALSE,message=FALSE}
wang.updated=merge(wang,TV0,by='mouse')
PGday.f = as.factor(wang.updated$PGday)
wang.updated=data.frame(wang.updated,PGday.f)
wang.updated$mouse=as.factor(wang.updated$mouse)
wang.updated$TV<-wang.updated$TV-wang.updated$TV0
logTV<-log(wang.updated$TV+1+abs(min(wang.updated$TV)),10)
wang.updated2<-data.frame(wang.updated,logTV)
require(car)
leveneTest(logTV~PGday.f*Treatment,data=wang.updated2,center=mean)
leveneTest(logTV~PGday.f*Treatment,data=wang.updated2)
wang.aov.rm=with(wang.updated2,aov(logTV~Treatment*PGday.f+Error(mouse/PGday.f)))
summary(wang.aov.rm)
wang.aov=with(wang.updated2,aov(logTV~Treatment*PGday.f))
summary(wang.aov)
TukeyHSD(wang.aov,"Treatment")
```

###Scatterplot of TV against PGday for each treatment
```{r}
wang.updated2=wang.updated2[order(wang.updated2$mouse,wang.updated2$PGday),]
wang.g1<-groupedData(TV~PGday | mouse, outer=~Treatment, order.groups=F,data=wang.updated2)
plot(wang.g1, display='mouse', outer=TRUE, aspect=2, key=F, xlab='PGday', ylab='TV', main='Individual TV data by Treatment Group')

wang.g1log<-groupedData(logTV~PGday | mouse, outer=~Treatment, data=wang.updated2)
plot(wang.g1log, display='mouse', outer=TRUE, aspect=2, key=F, xlab='PGday', ylab='log(TV)', main='Individual TV data by Treatment Group')
```

###**Build LMM models**
```{r,warning=F,message=F}
wang.glog<-groupedData(logTV~PGday|mouse, data=wang.updated2, order.groups=F)
wang.graw<-groupedData(TV~PGday|mouse, data=wang.updated2, order.groups=F)
```

####Compare model with or without sqaured terms, using raw TV
```{r,warning=F,message=F}
m7<-lme(TV~PGday*Treatment+I(PGday^2)+I(PGday^2):Treatment, random=~PGday+I(PGday^2)-1, method='ML', data=wang.graw)
m6<-lme(TV~PGday*Treatment, random=~PGday-1, method='ML', data=wang.graw)
anova(m6,m7)
```

####Build LMM model with raw TV and log transformed TV
```{r,warning=F,message=F}
wang.glog=with(data=wang.glog,wang.glog[order(Group,mouse,PGday),])
wang.graw=with(data=wang.graw,wang.graw[order(Group,mouse,PGday),])
m8<-lme(TV~PGday*Treatment+I(PGday^2)+I(PGday^2):Treatment, random=~PGday+I(PGday^2)-1, method='REML', data=wang.graw)
summary(m8)
m9<-lme(logTV~PGday*Treatment+I(PGday^2)+I(PGday^2):Treatment, random=~PGday+I(PGday^2)-1, method='REML', data=wang.glog)
```

####Fit mean profile curve for fixed effects
```{r,warning=F,message=F}
control = wang.updated2[wang.updated2$Treatment == 'Control',]
CTLA4=wang.updated2[wang.updated2$Treatment == 'hCTLA4',]
mAB2=wang.updated2[wang.updated2$Treatment == 'mAB2',]
par(mfrow=c(2,2))
curve(4.5*x^2-11.5*x+54.4,0,21,xlab="PGday",ylab="Marginal predicted TV",lty=3,lwd=2,ylim=c(-50,3200),co1=1,main="TV~PGday|Treatment curve")
curve(2.4*x^2-1.1*x+69.3,0,21,add=T,lty=3,lwd=2,col=2)
curve(1.6*x^2+4.4*x+50.5,0,21,add=T,lty=3,lwd=2,col=3)
legend("topleft",levels(wang.updated$Treatment),lty=3,lwd=2,col=1:3,bty='n')
curve(4.5*x^2-11.5*x+54.4,0,21,xlab="PGday",ylab="Marginal predicted TV",main="TV~PGday|Control",lty=3,lwd=3,ylim=c(-50,3200),co1=1)
points(control$PGday,control$TV,col='blue')
curve(2.4*x^2-1.1*x+69.3,0,21,ylim=c(-50,3200),xlab="PGday",ylab="Marginal predicted TV",main="TV~PGday|mAB2",lty=3,lwd=3,col=2)
points(mAB2$PGday,mAB2$TV,col='blue')
curve(1.6*x^2+4.4*x+50.5,0,21,ylim=c(-50,3200),xlab="PGday",ylab="Marginal predicted TV",main="TV~PGday|hCTLA4",lty=3,lwd=3,col=3)
points(CTLA4$PGday,CTLA4$TV,col='blue')
par(mfrow=c(1,1))
```

####Fit individual curve for each experimental subject
```{r,warning=F,message=F}
plot(augPred(m8,level=0:1),layout=c(5,3,1),main="Individual profile & Mean profile") #model bulit using Raw TV
plot(augPred(m9,level=0:1),layout=c(5,3,1),main="Individual profile & Mean profile(logTV)") #model bulit using log transformed TV
trellis.device(color=F)
```

####residue Diagnostics
```{r,warning=F,message=F}
plot(m8,resid(.,type="p")~fitted(.)|factor(Treatment),layout=c(3,1),aspect=2,abline=2)
plot(m8,resid(.)~PGday,abline=0)
plot(m9,resid(.,type="p")~fitted(.)|factor(Treatment),layout=c(3,1),aspect=2,abline=2)
qqnorm(m8,~resid(.)|factor(Treatment),layout=c(3,1),aspect=2,id=0.05)
```

####Diagnostics of random effects
```{r,warning=F,message=F}
qqnorm(m8,~ranef(.),id=0.1)#Check normality of random effects
pairs(m8,~ranef(.)|factor(Treatment),id=~mouse==8067,layout=c(3,1),aspect=2)
```

####Observed vs predicted value
```{r,warning=F,message=F}
plot(m8,TV~fitted(.)|factor(Treatment),id=0.05,layout=c(3,1),aspect=2,abline=c(0,1),lty=2)
```

####refit model after removing outlier 8067
```{r,warning=F,message=F}
wang.graw2 = wang.graw[wang.graw$mouse != 8067,]
m8.update = update(m8,data=wang.graw2)
summary(m8.update)
```

####Fit individual curve for each experimental subject and plot observed~fitted for updated model
```{r,warning=F,message=F}
plot(augPred(m8.update,level=0:1),layout=c(5,3,1),main="Individual profile & Mean profile")
plot(m8.update,TV~fitted(.)|factor(Treatment),id=0.05,layout=c(3,1),aspect=2,abline=c(0,1),lty=2)
```

##**Conclusion**
1.Two way RM ANOVA shows no significant treatment effect and no significant treatment*PGday interaction, while Two way ANOVA  
and Tukey's HSD shows there are significant difference between treatements and PGday, difference between mAB2-Control are significant.

2.LMM only shows significant main effect of PGday, but no significant treatment effect or treatment*PGday interaction.

3.Using module nlme seems to do a better job than lme4

4.using raw TV is better than log transformed TV, better agreement between fitted and observed values

5.one subject 8067 is the obvious outlier, after removing it, interaction between treatment and PGday^2 becomes significant













